---
title: "Formule 1 | Grand Prix de Monaco 2021 | Fernando Alonso"
author: "Baptiste Gorteau & Théo Cavenne"
output:
  html_document: default
  pdf_document: default
---

# Data description
```{r, include=FALSE}
# Import des packages :
library(fda)
#library(tidyfun)
library(viridis)
library(tidyverse)
library(RColorBrewer)
```

```{r echo=FALSE}
# Répertoire de travail :
#setwd("~/Desktop/M2 MAS/Signaux/Projet")
```

For this project, we decided to study Fernando Alonso's speed at the Monaco Formula 1 Grand Prix in 2021.

The datasets we will use are the following:

  - "laps.csv" : Summary with different information for each lap driven by Fernando Alonso (lap time, lap number, type of tires used, etc). The only variables we'll be using in this dataset are "LapNumber" (lap number), "PitOutTime" and "PitInTime" (times when Alonso entered or left the pits). We'll call this dataset "laps_alonso":
```{r echo=FALSE}
laps_alonso <- read.csv("Data/ALO/laps.csv")
print("laps_alonso")
print((laps_alonso %>% select(LapNumber,PitOutTime,PitInTime))[c(1,45,46),])
```
  - "lap_i.0.csv" (for i ranging from 1 to 77): Here we have 77 different data sets, each representing a lap of the Monaco GP 2021 driven by Fernando Alonso. For each of these data sets (laps), we'll have several Distance/Speed pairs representing the speed measured at a precise distance from the starting line for a lap. We'll concatenate these 77 datasets to produce a single dataframe (24861 rows and 3 columns) comprising the variables "Distance" (distance in meters from the starting line of the speed measurement point), "Speed" (speed in km/h measured) and "lap" (lap number (1 to 77)). We'll call this data set "ALO":
  
```{r echo=FALSE}
ALO <- read.csv(paste("Data/ALO/tours/lap_1.0.csv",sep = "")) %>% select(Distance,Speed) %>% mutate(lap=1)
for(i in 2:length(laps_alonso$LapNumber)){
  lap <- read.csv(paste("Data/ALO/tours/lap_",i,".0.csv",sep = "")) %>% select(Distance,Speed) %>% mutate(lap=i)
  ALO <- rbind(ALO,lap)
}
print("ALO")
print(head(ALO,10))
```

To facilitate our study, we have decided to remove the first lap and tire change laps from our dataset, as we feel they are not representative of a "normal" Grand Prix lap.

Indices of concerned laps :
```{r echo = FALSE}
delete_tour = sort(c(which(laps_alonso$PitOutTime!=""), which(laps_alonso$PitInTime!="")))
delete_tour
```


```{r echo=FALSE}
ALO <- ALO %>% filter(!(lap %in% delete_tour))
ALO$lap <- as.factor(ALO$lap)
```

## Visualization of the speed curve (as a function of distance from the starting line) for the 2nd lap:

```{r echo=FALSE}
lap2 <- ALO %>% filter(lap==2)
plot(lap2$Distance,lap2$Speed,pch=20,cex=0.5,xlab="Distance",ylab="Speed",main="Lap 2")
```

## Visualization of speed curves for all laps :

```{r echo=FALSE}
xrange = range(ALO$Distance)
yrange = range(ALO$Speed)
plot(seq(xrange[1],xrange[2],length=50),seq(yrange[1],yrange[2],length=50),type="n",main="Courbes observées",xlim=xrange,ylim=yrange,xlab="Distance",ylab="Speed")
for (i in unique(ALO$lap)){
  datap = ALO[which(ALO$lap==i),]
  lines(datap$Distance,datap$Speed,col=i)
}
```

The speed representations for each turn appear to be regular and non-periodic. We have therefore decided not to study the Fourier and wavelet bases. In the following, we shall concentrate on cubic spline bases.

# Choosing nodes for the cubic spline base

## First base

On the curve representation above, we can see that there are 24 "abrupt" changes (peaks) between acceleration and deceleration. We therefore decided to choose our nodes "by hand". The 24 nodes chosen represent the distances of the entrances and exits of these bends (relative to the starting line).

```{r echo=FALSE}
# Choix des noeuds
knots <- c(75,220,575,780,835,900,1020,1115,1180,1245,1290,1325,1360,1415,1900,2090,2260,2355,2565,2705,2785,2880,2930,2975)
print("Noeuds choisis :")
print(knots)
```

```{r echo=FALSE}
knots <- c(xrange[1],knots,xrange[2])
```
Here we have a 4th-order spline basis with 24 interior nodes and therefore 28 basis functions:

```{r echo=FALSE}
# Création de la base de splines avec les noeuds choisis :
bspbasis = create.bspline.basis(xrange,norder=4,breaks=knots)
```

 - Projection of individual curves onto the selected spline base :
 
```{r echo=FALSE}
laps = unique(ALO$lap)
nb_laps = length(laps)
CoeffMat = matrix(0,nrow=nb_laps,ncol=bspbasis$nbasis)
for (i in 1:nb_laps){
  decomp = Data2fd(ALO$Distance[ALO$lap==laps[i]],ALO$Speed[ALO$lap==laps[i]],basisobj = bspbasis)
  CoeffMat[i,] = as.vector(decomp$coefs)
}
head(CoeffMat)
```
 
 - Visualization of smoothed curves :

```{r echo=FALSE}
alo.smooth = fd(t(CoeffMat),bspbasis)  
plot(alo.smooth,xlab="Distance",ylab="Speed",main="Courbes lissées")
```
We note that 2 "abrupt" changes between acceleration and deceleration" are missing between 1100 and 1450, and 2 between 2275 and 3000.

## Second base

We'll reuse the same base, adding 5 extra nodes in each of these intervals (so adding 10 nodes in total).
In this case, we'll have a base of 4th-order splines with 34 interior nodes and therefore 38 basis functions.

 - Visualization of smoothed curves :

```{r echo=FALSE}
knots <- sort(c(knots,seq(1100,1450,length=5),seq(2775,3000,length=5)))

bspbasis = create.bspline.basis(xrange,norder=4,breaks=knots)

CoeffMat = matrix(0,nrow=nb_laps,ncol=bspbasis$nbasis)
for (i in 1:nb_laps){
  decomp = Data2fd(ALO$Distance[ALO$lap==laps[i]],ALO$Speed[ALO$lap==laps[i]],basisobj = bspbasis)
  CoeffMat[i,] = as.vector(decomp$coefs)
}
alo.smooth = fd(t(CoeffMat),bspbasis)  
plot(alo.smooth,xlab="Distance",ylab="Speed",main="Courbes lissées")
```
We find these representations very satisfactory. We shall therefore retain the spline base created above for the remainder of this section.

# Exploratory statistics

## Study of mean and variance

### Functional mean and standard deviation

```{r echo = FALSE}
plot(alo.smooth,col="gray",xlab="distance",ylab="speed", main = "Moyenne et variance des tours lissés")
fhat <- eval.fd(seq(xrange[1],xrange[2],length.out=325),alo.smooth)
mean_lap <- mean.fd(alo.smooth)
lines(mean_lap)
sd_lap <- sd.fd(alo.smooth)
lines(mean_lap+2*sd_lap,lwd=2,col=4,lty=2)
lines(mean_lap-2*sd_lap,lwd=2,col=4,lty=2)
legend("bottomleft",legend = c("Moyenne", "Écart-type"), col=c("black", "blue"), lty=1:2, cex=0.8)
```

The graph shows that most of the curves lie within the variance range. Greater variance can be observed at moments of change between acceleration and deceleration (local maximums and minimums). On the contrary, for the rest of the time, speed varies very little between laps. It is interesting to note that the variance at braking is higher when braking is strong (comparison between braking at 1700m and at 2500m).

## Covariance and correlation

### Covariance 

```{r echo = FALSE}
covspeed = var.fd(alo.smooth)
surfcov = eval.bifd(seq(min(ALO$Distance),max(ALO$Distance),length=325),seq(min(ALO$Distance),max(ALO$Distance),length=325),covspeed)
```

```{r echo = FALSE}
filled.contour(surfcov)
```

```{r echo = FALSE}
plot_ly(z = surfcov, type = "surface")
```

We have a problem that we can't solve. At values very close to zero, the covariance is extremely high. We have therefore decided to truncate our measurement intervals so that this peak is no longer visible.

```{r echo = FALSE}
plot_ly(z = surfcov[3:325,3:325], type = "surface")
```

The covariances shown here are rather low. However, there is a peak towards the end of the lap (bar 278). We can therefore assume that there is a strong variation between the curves at the end of the circuit. We can assume that this point corresponds to braking after 2.5km. Variations at this point can be seen in the variance representation above.

### Correlation

```{r echo = FALSE}
corspeed = cor.fd(seq(1,max(ALO$Distance),length=100),alo.smooth)
```

```{r echo = FALSE}
plot_ly(z = corspeed, type = "surface")
```

Thanks to this graph, we can clearly see that there is a strong correlation between points that follow each other. A point x and x+1 will be highly correlated. Which makes sense, given our data. This graph doesn't give us much information.

## Functional PCA

```{r echo = FALSE}
ACPF = pca.fd(alo.smooth,nharm=2,centerfns = TRUE)
ACPF$varprop
cumsum(ACPF$varprop)
```

Here, the first two principal components explain 54% of the data.

```{r echo = FALSE}
par(mfrow=c(1,2))
plot(mean_lap, main ="Moyenne des courbes lissées")
plot(ACPF$harmonics, main="Composantes principales")
legend("topright", legend = c("Composante 1", "Composante 2"), col=c("black", "red"), lty=1:2, cex=0.5)
par(mfrow=c(1,1))
```

The graph on the right shows the representations of the principal components of the PCA performed. The black curve is the one that captures the most information on the data. This same curve separates the data according to the first kilometer and the sector around 2.5km.

```{r echo = FALSE}
head(ACPF$scores)
plot(ACPF$scores[,1],ACPF$scores[,2],pch=3,xlab="FPC 1", ylab="FPC 2",type="n")
nomstours <- unique(ALO$lap)
type_pneu <- as.factor(laps_alonso$Compound)
text(ACPF$scores[,1],ACPF$scores[,2],labels=nomstours,cex=0.7,col=as.numeric(type_pneu))
```

This graph shows that axis 2 separates lap 48 from the other laps.
Let's take a look at which turn it was.

```{r echo = FALSE}
plot(mean_lap)
lines(ALO[ALO$lap==48,]$Distance,ALO[ALO$lap==48,]$Speed, col="red")
```

Indeed, on this lap, Ferando Alonso braked abnormally at the start of the lap (400m). Moreover, the peak of the second main axis corresponds to this distance of around 400m.

It's this braking that makes this tour different from the others.

If we return to the projection of laps on the axes, we can see that axis 1 separates laps on medium tires (black) from those on soft tires (red). 

If we go back to the representation of the variance and the analysis that was carried out, we can then think that the type of tire will have an influence on braking and acceleration. To try and confirm this hypothesis, we look at which types of tire are used when braking at 2.5km.  

```{r}
for(i in 1:nrow(laps_alonso))
if(laps_alonso$Compound[i]=="MEDIUM"){
  laps_alonso$color[i]="black"
} else if(laps_alonso$Compound[i]=="SOFT"){
  laps_alonso$color[i]="red"
}

plot(seq(xrange[1],xrange[2],length=50),seq(yrange[1],yrange[2],length=50),type="n",main="Courbes observées",xlab="Distance",ylab="Speed", xlim=c(2350, 2700),ylim=c(175,250))

for (i in unique(ALO$lap)){
  datap = ALO[which(ALO$lap==i),]
  lines(datap$Distance,datap$Speed,col=laps_alonso$color[which(laps_alonso$LapNumber==i)])
}
```

This graph shows that pre-braking speed is generally lower for laps on medium tires than on soft tires. This confirms our hypothesis.

From what we've been able to observe, the braking moments at 250m, 800m and 2.5km are surely the places on the circuit where Ferando Alonso can gain (or lose) the most time on his lap. The type of tire will also influence his braking at these points.